---
title: "Session 2 - Conditional branching and looping in R"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Intro_To_R_1Day/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


<p>&nbsp;</p>
These exercises are about the conditions and loops sections of [Introduction to R](https://rockefelleruniversity.github.io/RU_introtoR_abridged/). 

** Exercise 1 **

- Read in the file *categoriesAndExpression.txt* from the *data* directory of the course material. Filter it the *ofInterest* column to just the **Selected**.

```{r}
my_df <- read.table("data/categoriesAndExpression.txt",sep="\t",header=T)
head(my_df)
my_df_subset <- my_df[my_df$ofInterest == "Selected",]
my_df_subset
```

- Reorder the subset from smallest to largest in expression levels. 

```{r}
my_df_subset_ordered <- my_df_subset[order(my_df_subset$Expression),]
```

** Exercise 2 **

- Calculate the factorial (factorial of 3 = 3 * 2 * 1) of 10 using a loop.

```{r,echo=toMessage}

for(x in 1:10){
  if(x == 1){
    factorialAnswer <- 1
  }else{
    factorialAnswer <- factorialAnswer * x 
  }
}
factorialAnswer
```

- Using an ifelse() expression, create a factor from a vector of 1 to 40 where all numbers less than 10 are "small",10 to 30 are "mid",31 to 40 are "big"

```{r,echo=toMessage}
condExercise <- 1:40
condExercise
vectorResult <- ifelse(condExercise<10,"small",ifelse(condExercise < 31,"mid","big"))
temp <- factor(vectorResult,levels=c("small","mid","big"),order=T)
temp
```

** Exercise 3 **

- Read in all files from the *ExpressionResults* with .txt extension, using sapply and create a table of gene expression results.

NOTE: the *dir()* function can return only specific file types with the *pattern* argument. 

```{r,echo=toMessage}

filesToRead <- dir("ExpressionResults/", pattern = "*\\.txt", full.names=T)

fileRead <- sapply(filesToRead, read.delim, header=F, sep="\t")

mergedTable <- NULL
for(i in fileRead){
  if(is.null(mergedTable)){
    mergedTable <- i
  }else{
    mergedTable <- merge(mergedTable,i,by=1,all=T)
  }
  
  print(nrow(mergedTable))
}

mergedTable[1:3,] 

```


- Add annotation from *Annotation.txt* by merging it. How many of each Pathway are in the table? Can you also show this information with a plot?
 
```{r,echo=toMessage}
Annotation <- read.table("ExpressionResults/Annotation.txt",sep="\t",h=T)
annotatedExpression <- merge(Annotation,mergedTable,by=1,all.x=F,all.y=T)
annotatedExpression[1:2,]

annotatedExpression$Pathway <- factor(annotatedExpression$Pathway)


summary(annotatedExpression$Pathway)

plot(annotatedExpression$Pathway)

```
