---
title: "Session 1 - Data Types"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_introtoR_abridged/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

<p>&nbsp;</p>
These exercises are about the vector sections of [Introduction to R](https://rockefelleruniversity.github.io/RU_introtoR_abridged/). 



**Exercise 1**

- Create a vector containing the values 1,2,3,4,5
```{r,echo=toMessage}
x <- c(1,2,3,4,5)
x
```
- Create a vector containing the values 1 to 100
```{r,echo=toMessage}
x <- 1:100
x
```
- Create a vector containing the values  0,5,10,15,20
```{r,echo=toMessage}
x <- seq(0,20,5)
x
```
- Create a new vector, with all but the first and last value of the previous vector.
```{r,echo=toMessage}
y <- x[-c(1,length(x))]
y
```

**Exercise 2** 

- Create data frame called Annotation with a column of gene names ("Gene_1", "Gene_2", "Gene_3","Gene_4","Gene_5"),  ensembl gene names ("Ens001", "Ens003", "Ens006", "Ens007", "Ens010"),  pathway information ("Glycolysis", "TGFb", "Glycolysis", "TGFb", "Glycolysis") and  gene lengths (100, 3000, 200, 1000,1200).
```{r,echo=toMessage}
Annotation <- data.frame(geneNames=c("Gene_1", "Gene_2", "Gene_3","Gene_4","Gene_5"), ensembl=c("Ens001", "Ens003", "Ens006", "Ens007", "Ens010"),pathway=c("Glycolysis", "TGFb", "Glycolysis", "TGFb", "Glycolysis"),geneLengths=c(100, 3000, 200, 1000,1200))
```

- Add a column with the expression data: (1000, 3000, 10000,5000, 50) 
```{r,echo=toMessage}
Annotation$expression <- c(1000, 3000, 10000,5000,50) 
```

- Subset the data frame to only the first 2 genes and remove the ensembl IDs. 

```{r,echo=toMessage}
Annotation_subset <- Annotation[1:2,-2]
Annotation_subset
```

- Subset the data frame to the last 2 genes and remove the fourth and fifth column. Try do this so the code would work, whatever size the data frame is. 

```{r,echo=toMessage}
Annotation_subset2 <- Annotation[(nrow(Annotation)-1):nrow(Annotation),-c(4,5)]
Annotation_subset2
```

- Calculate the length normalized expression and add it to a new column

```{r,echo=toMessage}
Annotation$len_norm_exp <- Annotation$expression / Annotation$geneLengths
Annotation
```


