---
title: "Session 1 - Data Types"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_introtoR_abridged/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

<p>&nbsp;</p>
These exercises are about the vector sections of [Introduction to R](https://rockefelleruniversity.github.io/RU_introtoR_abridged/). 



**Exercise 1**

- Create a vector containing the values 1,2,3,4,5
```{r,echo=toMessage}
x <- c(1,2,3,4,5)
x
```
- Create a vector containing the values 1 to 100
```{r,echo=toMessage}
x <- 1:100
x
```
- Create a vector containing the values  0,5,10,15,20
```{r,echo=toMessage}
x <- seq(0,20,5)
x
```
- Create a new vector, with all but the first and last value of the previous vector.
```{r,echo=toMessage}
y <- x[-c(1,length(x))]
y
```

**Exercise 2** 

- Read in the file *categoriesAndExpression.txt* from the *data* directory of the course material.
```{r,echo=toMessage}
my_df <- read.table("data/categoriesAndExpression.txt",sep="\t",header=T)
```

- Subset the data frame to only the first 5 genes and remove the second column. 

```{r,echo=toMessage}
my_df_subset <- my_df[1:5,-2]
my_df_subset
```

- Subset the data frame to the last 5 genes and remove the second and third column. 

```{r,echo=toMessage}
my_df_subset <- my_df[(nrow(my_df)-4):nrow(my_df),-c(2,3)]
my_df_subset
```

- Calculate the log2 of the expression level for the subset data frame and add it as a new column.

```{r,echo=toMessage}
my_df_subset$log2_expression <- log2(my_df_subset$Expression)
my_df_subset
```

- Export your data as a excel spreadsheet
```{r,echo=toMessage}
library(rio)
export(my_df_subset,"data/subset_genes.xlsx")

```

